\documentclass{beamer}

\usepackage{color} % Package to include colors for syntax highlighting
\usepackage{graphicx}  % Package to include images
\usepackage{hyperref}  % Hyperlinks
\usepackage{tikz}

\setbeamertemplate{navigation symbols}{}  % Remove navigation symbols
\usecolortheme[light]{solarized}

\title{Mid module feedback: 2017-2018}
\date{}


\begin{document}

\frame{
    \titlepage
}

\frame{
    \Large
    \begin{quote}
        I don't think its necessary that when someone enters the room late the
        entire lecture is stopped to wait for that person
    \end{quote}
}

\frame{
    \Large
    \begin{quote}
        In lectures if no one answers a question after 30 seconds stop waiting
        :p
    \end{quote}

    \begin{quote}
        Don't like calling on people to answer questions because not everyone
        is comfortabl and confident enough to do so.
    \end{quote}
    % Emphasise that I'm very sympathetic to this and that I have not (and will
    % not) call on particular individuals (unless they raise their hand etc).
}

\frame{
    \Large
    \begin{quote}
        Computing is not a subject we can teach ourselves.
    \end{quote}
    \pause


    \vspace{2cm}
    ``Active learning increases student performance in science,
    engineering, and mathematics.'' \textit{Freeman et al.} 2014 (PNAS)
}

\frame{
    \begin{center}
        \Huge 
        \[> 0\]
    \end{center}
}

\frame{
    \Large
    \begin{quote}
        The labsheets are really well written, very detailed and the videos are
        really useful.
    \end{quote}
}

\frame{
    \Large
    \begin{quote}
        Work is very well structured in weekly patterns.
    \end{quote}
}

\frame{
    \Large
    \begin{quote}
        Lecture capture (although quality isn't great)
    \end{quote}
}

\frame{
    \Large
    \begin{quote}
        The review lecture on Thursday helps to clear up issues and highlight
        different ways of tackling a problem.
    \end{quote}
}

\frame{
    \Large
    \begin{quote}
        Vince is approachable
    \end{quote}
}

\frame{
    \Large
    \begin{center}
        \Huge 
        Constructive
    \end{center}
}


\frame{\frametitle{Class meetings}
    \Large
    \begin{quote}
        Material covered in lectures, finish early and not really planned
        sometimes
    \end{quote}

    \begin{quote}
        Your lectures are useful. They are well explained on what I find
        difficult.
    \end{quote}

    \begin{quote}
        Would prefer if we could go over the harder end questions in lectures,
        maybe start at the last question and work towards the first.
    \end{quote}
    % Excellent suggestion
}



\frame{\frametitle{Lab sessions}
    \Large
    \begin{quote}
        Lab sessions are sometimes unnecessary when you haven't completed the
        work.
    \end{quote}

    \begin{quote}
        Lab sessions are very helpful for when you have a problem with the code
        or are stuck on the lab sheet.
    \end{quote}

    \begin{quote}
        Only get help in lab session if indicated, maybe have PhD students go
        around class, interacting more.
    \end{quote}
}


\frame{\frametitle{Lab sheets}
    \Large
    \begin{quote}
        Help on how to solve problems such as the harder lab sheet questions
        would be appreciated (include a similar example not just for easier
        questions
    \end{quote}

    \begin{quote}
        Sometimes I wish we had longer than a week to finish a lab sheet.
    \end{quote}

    \begin{quote}
        The pace...
    \end{quote}
    % Explain that I believe to have done this but completely understand. The
    % debuging question and the worked example question are always examples very
    % similar to the first 1 or two other questions.
}



\frame{

    \begin{center}
        \Huge
        Questions
    \end{center}
}

\frame{

    \Large
    \begin{quote}
        Including recommended reading/textbooks?
    \end{quote}
    % Point students at list of recommended reading/textbooks
}


\frame{
    \Large
    \begin{quote}
        Less lectures and more labs to actually get feedback as opposed to
        just watching.
    \end{quote}
    % Love this comment: this student gets it.
    % Will discuss 


    \begin{quote}
        Additional (optional) lab sessions would be good
    \end{quote}

}

\frame{
    \Large
    \begin{quote}
        Very difficult to catch up if you don't understand something earlier
        which is used later.
    \end{quote}
    % Agree! Clarify that because of the continuous feedback this is made
    % clearer. However that this is the case with all subjects.
}

\frame{
    \begin{quote}
        I sometimes find it difficult to interpret how a function or library
        is supposed to work and how to use it properly when learning about them
        through online research. I get why its useful to do the research and how
        that could develop our skills but it might help people if there was a
        demo in the solutions (or something).
    \end{quote}
    % This comment puzzled me: I plan to discuss it in class and attempt to
    % clarify, perhaps my suggested highlighting will make things clearer.
}

\frame{
    \begin{center}
        \Huge
        Other
    \end{center}
}

\frame{
    \Large
    \begin{quote}
        I don't like class at 9am
    \end{quote}
}


\frame{
    \Large
    \begin{quote}
        Some sort of club?
    \end{quote}
}

\frame{
    \Large
    \begin{quote}
        Relax a little you scare me
    \end{quote}
}

\frame{
    \Large
    \begin{quote}
        Where are you from?
    \end{quote}
}

\frame{
    \Large
    \begin{quote}
        Vince is alright
    \end{quote}
}

\frame{
    \begin{center}
        \Huge 
        Finally.
    \end{center}
}

\frame{
    \Large
    \begin{quote}
        I don't know where to get help outside of the labs.
    \end{quote}
}


\frame{
    \Large
    \begin{quote}
        More clarity of end game eg what happens with course/examination etc.
    \end{quote}
}

\end{document}
